Characteristic polynomials of p-adic matrices

### 2017-02-03 charpoly/charpoly.tex
Optimal precision on eigenvalues

@@

\newcommand{\inv}{\text{\rm inv}}

@@

Let us consider more precisely the case where $M$ is given at 
jagged precision: the $(i,j)$ entry of $M$ is given at precision 
$O(\pi^{N_{i,j}})$ for some given integer $N_{i,j}$. 
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then teaches us that
the optimal precision on the $k$-th coefficient of $\chi_M$ is 
$O(\pi^{M_k})$ where $M_k$ is given by the formula:
\begin{equation}
\label{eq:optjagged}
M_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
\end{equation}
where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.

@@

\begin{rem}
If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
then we can avoid the final base change step in the Hessenberg method.
Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
we can write:
$$tr(\com (X{-}M) \cdot dM)=tr(\com (X{-}H)\cdot P^{-1} dM P)$$
where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
As a consequence, the integer $M_k$ given the optimal precision on the 
$k$-th coefficient of $M$ is also equal to
$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
where $C^H_{i,j}$ is the $(i,j)$ entry of the comatrix of $X{-}H$
where $H$ is the Hessenberg form of $M$.
\end{rem}

@@

\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item the column ``Optimal'' is the average of the quantities 
$(M_k{-}v) - N$ (where $M_k$ is defined by Eq.~\eqref{eq:optjagged}): 
$M_k{-}v$ is the optimal \emph{relative} precision, so that the
difference $(M_k{-}v) - N$ is the loss of relative precision;
\item the column ``CR'' is the average of the quatities 
$(\text{CR}_k{-}v) - N$ where $\text{CR}_k$ is the computed (absolute) 
precision on the $k$-th coefficient of $\chi_M$;
\item the column ``FP'' is the average of the quatities 
$(\text{FP}_k{-}v) - N$ where $\text{FP}_k$ is the first position of
an incorrect digit on the $k$-th coefficient of $\chi_M$.
\end{enumerate}

@@

\todo{Optimal precision on $P(\lambda_i)$ and on $\lambda_i$}

\subsection{Legendre precision}

\todo{Describe Legendre precision and show that it is optimal for
a diagonalizable matrix over $\Zp$.}

%\subsection{Algorithm}

>>>>

\newcommand{\calU}{\mathcal{U}}

@@

\newcommand{\inv}{\text{\rm inv}}
\newcommand{\app}{\text{\rm app}}

@@

Let us consider more precisely the case where $M$ is given at 
jagged precision: the $(i,j)$ entry of $M$ is given at precision 
$O(\pi^{N_{i,j}})$ for some given integer $N_{i,j}$. 
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then teaches us that
the optimal precision on the $k$-th coefficient of $\chi_M$ is 
$O(\pi^{N'_k})$ where $N'_k$ is given by the formula:
\begin{equation}
\label{eq:optjagged}
N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
\end{equation}
where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.

@@

\begin{rem}
If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
then we can avoid the final base change step in the Hessenberg method.
Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
we can write:
$$\tr(\com (X{-}M) \cdot dM)=\tr(\com (X{-}H)\cdot P^{-1} dM P)$$
where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
As a consequence, the integer $N'_k$ given the optimal precision on the 
$k$-th coefficient of $M$ is also equal to
$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
where $C^H_{i,j}$ is the $(i,j)$ entry of the comatrix of $X{-}H$
where $H$ is the Hessenberg form of $M$.
\end{rem}

@@

\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item the column ``Optimal'' is the average of the quantities 
$(N'_k{-}v) - N$ (where $N'_k$ is defined by Eq.~\eqref{eq:optjagged}): 
$N'_k{-}v$ is the optimal \emph{relative} precision, so that the
difference $(N'_k{-}v) - N$ is the loss of relative precision;
\item the column ``CR'' is the average of the quatities 
$(\text{CR}_k{-}v) - N$ where $\text{CR}_k$ is the computed (absolute) 
precision on the $k$-th coefficient of $\chi_M$;
\item the column ``FP'' is the average of the quatities 
$(\text{FP}_k{-}v) - N$ where $\text{FP}_k$ is the first position of
an incorrect digit on the $k$-th coefficient of $\chi_M$.
\end{enumerate}

@@

Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple} eigenvalue 
of $M$. We are interesting in quantifying the optimal precision on 
$\lambda$ when $M$ is given with some uncertainty.

For doing so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
suppose that the uncertainty of $M$ is ``jagged'' in the sense that
each entry of $M$ is given at some precision $O(\pi^{N_{i,j}})$.
Let $\lambda_\app$ be the relevant eigenvalue of $M_\app$. We remark 
that it is possible to follow the eigenvalue $\lambda_\app$ on a small 
neighborhood $\calU$ of $M$. Precisely, there exists a unique continuous 
function $f : \calU \to K$ such that:
\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item $f(M_\app) = \lambda_\app$, and
\item $f(M)$ is an eigenvalue of $M$ for all $M \in \calU$.
\end{enumerate}

\begin{lem}
The function $f$ is strictly differentiable on a neighborhood of 
$M_\app$.
The differential of $f$ at $M$ is the linear mapping:
$$dM \mapsto d \lambda = - \frac{\tr(\com(\lambda-M) \cdot dM)}
{\chi'_M(\lambda)}$$
where $\chi'_M$ is the usual derivative of $\chi_M$.
\end{lem}

\begin{proof}
The first assertion follows from the implicit function Theorem.
Differentiating the relation $\chi_M(\lambda) = 0$, we get
$\chi'_M(\lambda) \cdot d \lambda + \tr(\com(X-M) \cdot dM)(\lambda) = 0$,
from what the Lemma follows.
\end{proof}

\noindent
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} now implies that, if the
$N_{i,j}$'s are large enough and rather well balanced, the optimal
precision on the eigenvalue $\lambda$ is $O(p^{N'})$ with:
$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(C_{i,j}(\lambda)) - 
\val(\chi'_M(\lambda))\big)$$
where $C_{i,j}$ denotes as above the $(i,j)$ entry of $\com(X{-}M)$.
Writing
$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t \text{ mod } \chi_M$
as in Proposition~\ref{prop:shortcom}, we find:
\begin{align*}
N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big)
\end{align*}
where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
denotes the $j$-th row of $Q$. Note moreover that $V(\lambda)$ is
the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
By the discussion of \S \ref{}, the exact value of $M$ can be 
determined for a cost of $\softO(n^\omega)$ operations in $K$ and
$O(n^2)$ operations on integers.

When $M$ is given at flat precision, \emph{i.e.} the $N_{i,j}$'s are all 
equal to some $N$, the formula for $N'$ rewrites:
\begin{align*}
N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
+ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big)
\end{align*}
and can therefore now be evaluated for a cost of $\softO(n^\omega)$
operations in $K$ and only $O(n)$ operations on integers.

To conclude with, let us discuss briefly the situation where we want
to figure out the optimal jagged precision on a tuple $(\lambda_1,
\ldots, \lambda_s)$ of eigenvalues. Applying what we have done
previously, we find that the optimal precision on $\lambda_k$ is 
\begin{align*}
N'_k & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big).
\end{align*}

\begin{prop}
\label{prop:opteigenvalues}
The $N'_k$'s can be all computed in $\softO(n^\omega)$ operations
in $K$ and $O(n^2 s)$ operations on integers.

If the $N_{i,j}$'s are all equal, the above complexity can be 
lowered to $\softO(n^\omega)$ operations in $K$ and $O(n s)$ operations 
on integers.
\end{prop}

\begin{proof}
The $\alpha(\lambda_k)$'s and the $\chi'_M(\lambda_k)$'s can be computed 
for a cost of $\softO(ns)$ operations in $K$ using fast multipoint 
evaluation methods~\cite{}.
On the other hand, we observe that $P_i V(\lambda_k)^t$ is nothing but
the $(i,k)$ entry of the matrix:
$$P \cdot \left( \begin{matrix}
\lambda_1 & \cdots & \lambda_s \\
\lambda_1^2 & \cdots & \lambda_s^2 \\
\vdots & & \vdots \\
\lambda_1^{n-1} & \cdots & \lambda_s^{n-1}
\end{matrix} \right).$$
The latter product can be computed in $\softO(n^\omega)$ operations in 
$K$\footnote{It turns out that $\softO(n^2)$ is also possible because
the right factor is a structured matrix (a truncated Vandermonde):
computing the above product reduces to evaluate a polynomial at the
points $\lambda_1, \ldots, \lambda_s$.}. Therefore all the $P_i 
V(\lambda_k)^t$'s (for $i$ and $k$ varying) 
can be determined with the same complexity. 
Similarly all the $V(\lambda) Q_j^t$ are computed for the same cost.
The first assertion of Proposition~\ref{prop:opteigenvalues} follows.
The second assertion is now proved similarly to the case of a unique
eigenvalue.
\end{proof}
